---
title: "Modelling and Plotting Count Data"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

```{r}
library("OnRampsR")
library("tidyverse")
library("broom")
library("readxl")
library("lubridate")
library("knitr")
library("XLConnect")
```

This project tests if the sex of a head of state/government had an effect on its Covid-19 outcomes.
Background: A female [twitter user respnded](https://twitter.com/eawilloughby/status/1249852024067416064) to a [Forbes article that went viral while claiming that countries with woemn leaders had better Covid-19 outcomes](https://www.forbes.com/sites/avivahwittenbergcox/2020/04/13/what-do-countries-with-the-best-coronavirus-reponses-have-in-common-women-leaders/).
I was concerned that the conclusion of "no effect" might be hasty because the twitter user made two mistakes:
1. She applied the incorrect statistical test (count data should be tested with a Poisson distribution not a t-test) 
2. She also drastically truncated the data (see the discrepancy between her t-tests' degrees of freedom, t(df), and the N's)

Therefore, I tested the full dateset (N = 213) with a model specific to count data (models for non-normal data are called a generalized linear models).

Lastly, simple numbers of cases & deaths are incomparable between countries because the onset of the pandemic happened on different dates. (The twitter user above should have simply compared case fatality rates.) To compare cases and deaths across countries don't compare curves of per capita rates, but curves of per capita rates starting at the same reference point [(example, the day since each country had 1death/million people)](https://twitter.com/jonfwilkins/status/1249511124749422593).



```{r warning = FALSE}
clean <- read_csv("Modelling and Plotting Count Data.csv")

names(clean) <- c("country", "cases", "deaths", "cases/million", "deaths/million", "leader_name",
                  "designation", "sex")

# glm of cfr (ie deaths/cases) predicted by sex; applying poisson log transforms the outcome variable

# 1. testing if the two levels differ from each other
full_0 <- glm((deaths/cases) ~ sex, family = poisson(link = log), data = clean) %>% 
car::Anova(type = 3) %>% tidy() ##chi-squared test for difference between two levels of the factor sex

# 2. testing if the two levels differ from zero (for calculation of CIs)
full <- glm((deaths/cases) ~ 0 + sex, family = poisson(link = log), data = clean) %>% tidy(conf.int = T)

# renaming the contents of the first coln of the tidy output to make plotting easier
full$term <- full$term %>% as.factor() # first encode the two rows of the first coln as factor levels
levels(full$term) <- c("Men", "Women") #change their names from "sexMen" & "sexWomen" to "Men" & "Women"

#plotting
ggplot(full, aes(term, estimate)) +  theme_classic() + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high), #makes gray boxes equal to length of CI
                     size = 22, alpha = 0.1) + #determines width and fading of boxes
  geom_boxplot(aes(y = estimate), width = 0.07) + #marks line at average values within boxes
  geom_point(data = clean, aes(x = sex, y = log(deaths/cases)), #overlays data points
                         position = position_jitter(0.05), alpha = 0.4) + #jitters and fades data points
    theme(plot.title = element_text(vjust = - 10, hjust = 0.5), # shifts title 
          plot.subtitle = element_text(vjust = - 11, hjust = 0.5))  +  # shifts subtitle 
  labs(title="Generalized linear model of CFR vs Leader's sex", #title
         subtitle = "(Chi-sq for sex diff = 0.008, p-value = 0.930)", #subtitle
         x="Leader's sex", # x-axis title
         y = "log(Case fatality rate)") + #y-axis title
  scale_y_continuous(limits =  c(-7, -1)) #increases the scale of the y-axis

```