---
title: "Untitled"
author: "Ameek Bhalla"
date: "10/31/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Agenda
## Life History
## Genetic models of different diseases
[] rebuild Berg and Geschwind figure
## Open science
[] Science ambassador certificate
## Agenda Item #1
## The problem of missing heritability
## Controvery around solutions to missing heritability
## Limitations of previous studies
## PhD schematic
[] Write out model equation
## Analysis pipeline
## Analysitcal challenges
## How to treat sex
[] sensitivity analysis for effect of sex
## Quantile normalization -- within or across
[] simulation for effect of normalization
## Can we estimate random effects
[] semnsitivity analysis of random effects
## My soltuions have general applicability
[] how much did the overall power increase
## PhD results
## Summary side
## Domian specific knowledge
## Match between Sage ethos and mine

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```



